Weighted Model Counting Without Parameter Variables

### 2021-02-28 doc/paper3/paper.tex
Lots of writing

\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage[binary-units]{siunitx}

@@

@@

% TODO: we also show correctness
% TODO: and new notation!

@@

NOTE: We list some properties of the operations on pseudo-Boolean functions
discussed in this section that can be conveniently represented using our syntax.
The proofs of all these properties follow directly from the definitions.

@@

\paragraph{Notes.}
\begin{itemize}
\item Weights on literals other than the positive literals that correspond to
  variables in $X_P$ are redundant as they either are equal to one or duplicate
  an already-defined weight.
\item Mention that formulas, clauses, and models are all treated as sets.
\end{itemize}

NOTE: Our definition of WMC is largely based on the standard definition
\cite{DBLP:journals/ai/ChaviraD08}, but explicitly partitions variables into
indicator and parameter variables.

\begin{definition}[WMC Instance]
  A \emph{WMC instance} is a tuple $(\phi, X_I, X_P, w)$, where $X_I$ is
  the set of indicator variables, $X_P$ is the set of parameter variables (with
  $X_I \cap X_P = \emptyset$), $\phi$ is a propositional formula in CNF over
  $X_I \cup X_P$, and $w\colon X_I \cup X_P \cup \{\neg x \mid x \in X_I \cup
  X_P\} \to \mathbb{R}$ is the weight function. The \emph{answer} of the
  instance is $\sum_{Y \models \phi} \prod_{Y \models l} w(l)$.
\end{definition}

@@

NOTE: Adding scaling factor $\omega$ to the definition allows us to remove
clauses that consist entirely of a single parameter variable. The idea of
extracting some of the structure of the WMC instance into an external
multiplicative factor originates in one of the Bayesian network encodings
\cite{DBLP:conf/ecai/BartKLM16}.

\subsection{Bayesian Network Encodings}

\paragraph{TODO}
\begin{itemize}
\item Define the terminology surrounding BNs: variables, values, directed
  probabilistic graphical model, DAG, binary = two values per variable.
\item Reuse this term where I'm referring to two values.
\item Give a brief description of each encoding.
\item Talk about the parameter/indicator variable distinction.
\end{itemize}

WMC is a well-established technique for Bayesian network inference, particularly
effective on networks where most variables have only a few possible values
\cite{DBLP:conf/kr/Darwiche02}. Many ways of encoding a Bayesian network into a
WMC instance have been proposed. We will refer to them based on initials of the
authors and the year of publication: \texttt{d02}
\cite{DBLP:conf/kr/Darwiche02}, \texttt{sbk05} \cite{DBLP:conf/aaai/SangBK05},
\texttt{cd05} \cite{DBLP:conf/ijcai/ChaviraD05}, \texttt{cd06}
\cite{DBLP:conf/sat/ChaviraD06}, \texttt{bklm16} \cite{DBLP:conf/ecai/BartKLM16}.

\begin{remark}
  Strictly speaking, encodings such as \texttt{cd05} and \texttt{cd06} are
  \emph{not} WMC encodings. Instead, they encode Bayesian network inference into
  instances of the \emph{minimum-cardinality} WMC problem, where the answer is
  defined to be $\sum_{Y \models \phi\text{, }|Y| = k} \prod_{Y \models l}
  w(l)$, where $k = \min_{Y \models \phi\text{, }Y \ne \emptyset} |Y|$, if $k$
  exists, otherwise the answer is zero. This additional condition on model
  cardinality becomes necessary because these encodings eliminate clauses of the
  form $p \Rightarrow i$, where $p \in X_P$ is a parameter variable, and $i \in
  X_I$ is an indicator variable. Nonetheless, our transformation algorithm still
  works on such encodings, although the experimental results are discouraging
  because they use approximately twice as many indicator variables. For
  instance, each binary variable of a Bayesian network is encoded using two
  indicator variables while one would suffice.
\end{remark}

\section{Parameter Variable Elimination}

\paragraph{Notes.}
\begin{itemize}
\item Let $X_P$ be the set of parameter variable and $X_I$ be the set of
  indicator variables.
\item Parameter variables are either taken from the LMAP file (for encodings
  produced by Ace) or assumed to be the variables that have both weights equal
  to 1.
\item If a parameter variable in a clause is `negated', we can ignore the
  clause. We assume that there are no clauses with more than one instance of
  parameter variables.
\item The second \textbf{foreach} loop can be performed in constant time by
  representing $\phi'$ as a list and assuming that the two 'clauses' are
  adjacent in that list (and incorporating it into the first loop).
\item The $d$ map is constructed in $\mathcal{O}(|X_P|\log|X_P|)$ time (we want
  to use a data structure based on binary search trees rather than hashing).
\item \texttt{rename} can be implemented in $\mathcal{O}(\log |X_P|)$ time.
\item This may look like preprocessing, but all the transformations are local
  and thus can be incorporated into an encoding algorithm with no slowdown. In
  fact, if anything, the resulting algorithm would be slightly faster, as it
  would have less data to output.
\item We do this as a follow-up to Ace compilation because the implementations
  of the encodings are all closed-source, and we don't want to re-invent (a
  subpar version of) the wheel.
\item No obvious way to do this for sbk05 because the roles of indicator and
  parameter (i.e., `chance') variables overlap.
\item This just says that $p$ is equivalent to a conjunction.
\item The first group of conditions applies to \texttt{d02}, while the second
  group applies to \texttt{bklm16}.
\item For \texttt{cd05} and \texttt{cd06}, condition $2bi$ should be replaced
  with $w(\neg p) = 1$.
\item Benefits of having this proof in the paper:
  \begin{itemize}
  \item It puts all encodings on a common ground.
  \item It illustrates the convenience of our notation for reasoning about
    (certain types of) pseudo-Boolean functions.
  \item It's too big and too important to be left for the appendix.
  \end{itemize}
\end{itemize}

\begin{algorithm}[t]
  \caption{WMC instance transformation}
  \label{alg:transformation}
  \SetKwFunction{rename}{rename}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{WMC instance $(\phi, X_I, X_P, w)$}
  \KwResult{PBP instance $(F, X, \omega)$}
  $F \gets \emptyset$\;
  $\omega \gets 1$\;
  let $d\colon X_P \to \mathbb{N}$ be defined as $p \mapsto |\{ o \in X_P \mid o
  \le p \}|$\;
  \ForEach{clause $c \in \phi$\label{line:foreach1start}}{
    \uIf{$c \cap X_P = \{ p \}$ for some $p$ \textnormal{\textbf{and}} $w(p) \ne
      1$}{
      \uIf{$|c| = 1$}{$\omega \gets \omega \times w(p)$\;}
      \Else{
        $F \gets F \cup \left\{ \left[ \bigwedge_{l \in c \setminus
              \{p\}} \neg l \right]^{w(p)}_1 \right\}$\;
      }
    }
    \ElseIf{$\{p \mid \neg p \in c\} \cap X_P  = \emptyset$}{
      $F \gets F \cup \{ [c]^1_0 \}$\; \label{line:foreach1end}
    }
  }
  \ForEach{indicator variable $v \in X_I$ \label{line:foreach2start}}{
    \If{$\{[v]_1^p, [\neg v]_1^q\} \subseteq F$ for some $p$ and $q$}{
      $F \gets F \setminus \{ [v]_1^p, [\neg v]_1^q \} \cup \{ [v]_q^p
      \}$\; \label{line:foreach2end}
    }
  }
  replace every variable $v$ in $F$ with $\rename{$v$}$\;
  \Return{$(F, X_I, \omega)$}\;
  \Fn{\rename{$v$}}{
    $S \gets \{u \in X_P \mid u \le v\}$\;
    \lIf{$S = \emptyset$}{\Return{$v$}}
    \Return{$v - d(\max S)$}\;
  }
\end{algorithm}

@@

@@

% NOTE: mention that this is similar (but not equal) to the DPMC/ADDMC result
\begin{proposition} \label{prop:equivalence}
  Let $(\phi, X_I, X_P, w)$ be a WMC instance. Then
  \begin{equation}
  \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
        x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
    1\right) \label{eq:new_wmc}
  \end{equation}
  is a PBP instance with the same answer.
\end{proposition}
\begin{proof} % TODO: equation is not the right word
  Let $f = \prod_{c \in \phi} [c]_0^1$, and $g = \prod_{x \in X_I \cup X_P}
  [x]_{w(\neg x)}^{w(x)}$. Then the answer of \cref{eq:new_wmc} is
  \[
    (\exists_{X_I \cup X_P} fg)(\emptyset) = \sum_{Y \subseteq X_I \cup X_P}
    (fg)(Y) = \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y)
  \]
  by \cref{lemma:sum}. Note that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that
  \[
    \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models \phi} \prod_{Y
      \models l} w(l)
  \]
  as required.
\end{proof}

\begin{theorem}[Correctness]
  \Cref{alg:transformation}, when given a WMC instance $(\phi, X_I, X_P,
  w)$, returns PBP instance with the same answer, provided the following
  conditions are satisfied:
  \begin{enumerate}
  \item for all indicator variables $i \in X_I$, $w(i) = w(\neg i) = 1$,
  \item and either
    \begin{enumerate}
    \item for all parameter variables $p \in X_P$, there is a non-empty family
      of literals $(l_i)_{i=1}^n$ such that \label{condition:d02}
      \begin{enumerate}
      \item $w(\neg p) = 1$,
      \item $l_i \in X_I$ or $\neg l_i \in X_I$ for all $i = 1, \dots, n$,
        \label{condition:equivalence1}
      \item and $\{ c \in \phi \mid p \in c \text{ or } \neg p \in c \} =
        \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
        \mid i = 1, \dots, n \}$; \label{condition:equivalence2}
      \end{enumerate}
    \item or for all parameter variables $p \in X_P$, \label{condition:bklm16}
      \begin{enumerate}
      \item $w(p) + w(\neg p) = 1$,
      \item for any clause $c \in \phi$, $|c \cap X_P| \le 1$,
      \item there is no clause $c \in \phi$ such that $\neg p \in c$,
      \item if $\{ p \} \in \phi$, then there is no clause $c \in \phi$ such
        that $c \ne \{ p \}$ and $p \in c$, \label{cond:just_parameter}
        % 'prime implicants are logically disjoint'
      \item and for any $c, d \in \phi$ such that $c \ne d$, $p \in c$ and $p
        \in d$, $\bigwedge_{l \in c \setminus \{ p \}} \neg l \land \bigwedge_{l
          \in d \setminus \{ p \}} \neg l$ is false. \label{cond:disjoint}
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}
\end{theorem}
\begin{proof}
  By \cref{prop:equivalence},
  \begin{equation}
    \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
          x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
      1\right) \label{eq:new_wmc2}
  \end{equation}
  is a PBP instance with the same answer as the given WMC instance. By
  \cref{def:new_wmc}, its answer is
  \begin{equation} \label{eq:answer}
    \left(\exists_{X_I \cup X_P} \left( \prod_{c \in \phi} [c]_0^1 \right) \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset)
  \end{equation}
  Since both \cref{condition:d02,condition:bklm16} ensure that each clause in
  $\phi$ has at most one parameter variable, we can partition $\phi$ into
  $\phi_* \coloneqq \{c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \emptyset \}$
  and $\phi_p \coloneqq \{ c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \{ p \}
  \}$ for all $p \in X_P$. We can then use \cref{thm:early} to reorder
  \eqref{eq:answer} into
  \[
    \left(\exists_{X_I} \left( \prod_{x \in X_I} [x]_{w(\neg x)}^{w(x)} \right)
      \left( \prod_{c \in \phi_*} [c]_0^1 \right) \prod_{p \in X_P} \exists_p
      [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 \right)(\emptyset).
  \]

@@

\section{Experimental Results}

\paragraph{Notes.}
\begin{itemize}
\item The experiments were run on a computing cluster with Intel Xeon E5-2630,
  Intel Xeon E7-4820, and Intel Xeon Gold 6138 processors with a
  \SI{1000}{\second} timeout separately on both encoding and inference, and a
  \SI{32}{\gibi\byte} memory limit.\footnote{Each instance was run on the same
    processor across all algorithms and encodings.}
\item Along with
  \textsf{DPMC}\footnote{\url{https://github.com/vardigroup/DPMC}}, in our
  experiments we also include WMC algorithms used in the papers that introduce
  each encoding: \textsf{Ace}
  for \texttt{cd05}, \texttt{cd06}, and \texttt{d02};
  \textsf{Cachet}\footnote{\url{https://cs.rochester.edu/u/kautz/Cachet/}}
  \cite{DBLP:conf/sat/SangBBKP04} for \texttt{sbk05}; and
  \textsf{c2d}\footnote{\url{http://reasoning.cs.ucla.edu/c2d/}}
  \cite{DBLP:conf/ecai/Darwiche04} with
  \textsf{query-dnnf}\footnote{\url{http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html}}
  for \texttt{bklm16}.
\item \textsf{DPMC} was run in \emph{greedy mode}. This mode (which was not part
  of the original paper \cite{DBLP:conf/cp/DudekPV20}) optimises the order in
  which pseudo-Boolean functions are multiplied by prioritising functions with
  small representations.
\item Indeed, the numbers of variables in \texttt{bklm16++} and \texttt{d02++}
  encodings are equal whenever all variables in a Bayesian network have two
  values (which applies to most of the benchmark instances). Nonetheless,
  \texttt{bklm16++} performs much faster than \texttt{d02++}.
\item \textsf{DPMC} is run with tree decomposition based planning and algebraic
  decision diagram based execution---the best-performing combination in the
  original set of experiments \cite{DBLP:conf/cp/DudekPV20}. We use a single
  iteration of \textsf{htd} \cite{DBLP:conf/cpaior/AbseherMW17} to generate
  approximately optimal tree decompositions---we found that this configuration
  is efficient enough to handle huge instances, and yet the width of the
  returned decomposition is unlikely to differ from optimal by more than one or
  two.
\item We also record numbers of variables in each encoding before and after our
  additional transformation.
\item Experimental research questions:
  \begin{itemize}
  \item Can parameter variable elimination improve inference speed?
  \item How does DPMC combined with encodings without (and with) parameter
    variables compare with other WMC algorithms and other encodings?
  \item Which instances is our approach particularly successful on (compared to
    other algorithms and encodings and to the same encoding before our
    transformation)?
  \item What proportion of variables is typically removed?
  \item Do some encodings benefit from this transformation more than others?
  \end{itemize}
\item TODO: How many times am I faster than old cd06 and new bklm16? Perhaps add
  other numbers as well.
\item To perform Bayesian network inference with \textsf{DPMC} (or with any
  other WMC algorithm not based on compilation such as \textsf{Cachet}), one
  needs to select a probability to compute
  \cite{DBLP:conf/cp/DudekPV20,DBLP:conf/sat/SangBBKP04}. If a network comes
  with an evidence file, we compute the probability of this evidence. Otherwise,
  let $X$ be the variable last mentioned in the Bayesian network file. If
  \textsf{true} is one of the values of $X$, then we compute $\Pr(X =
  \textsf{true})$, otherwise we choose the first-mentioned value of $X$.
\end{itemize}
% TODO: describe the results.

>>>>

\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage[binary-units]{siunitx}
\usepackage{booktabs}
\usepackage{capt-of}

@@

% TODO for later:
% 1) read through v2 of the paper and check if I forgot to mention something
% here
% 2) run this through grammarly

@@

\paragraph{Notes.}
\begin{itemize}
\item we also show correctness
\item and introduce a new notation
\end{itemize}

TODO: Mention that formulas, clauses, and models are all treated as sets.

\begin{definition}[WMC]
  A \emph{WMC instance} is a tuple $(\phi, X_I, X_P, w)$, where $X_I$ is
  the set of indicator variables, $X_P$ is the set of parameter variables (with
  $X_I \cap X_P = \emptyset$), $\phi$ is a propositional formula in CNF over
  $X_I \cup X_P$, and $w\colon X_I \cup X_P \cup \{\neg x \mid x \in X_I \cup
  X_P\} \to \mathbb{R}$ is the weight function. The \emph{answer} of the
  instance is $\sum_{Y \models \phi} \prod_{Y \models l} w(l)$.
\end{definition}

\begin{definition}[Minimum-Cardinality WMC]
  A \emph{minimum-cardinality WMC} instance has same tuple as a WMC instance,
  but its answer is defined to be $\sum_{Y \models \phi\text{, }|Y| = k}
  \prod_{Y \models l} w(l)$, where $k = \min_{Y \models \phi\text{, }Y \ne
    \emptyset} |Y|$, if $k$ exists, otherwise the answer is zero.
\end{definition}

Our definition of WMC is largely based on the standard definition
\cite{DBLP:journals/ai/ChaviraD08}, but explicitly partitions variables into
indicator and parameter variables. In practice, we identify this partition in
one of two ways. If an encoding is generated by \textsf{Ace}, then variable
types are explicitly identified in the LMAP file generated alongside the
encoding. Otherwise, we define a variable $x$ to be a parameter variable if
$w(x) = w(\neg x) = 1$.

\subsection{Bayesian Network Encodings (mostly done)}

A \emph{Bayesian network} is a directed acyclic graph with random variables as
vertices and edges as conditional dependencies. As is common in related
literature
\cite{DBLP:conf/kr/Darwiche02,DBLP:conf/aaai/SangBK05}, we assume that each
variable has a finite number of values. We call a Bayesian network \emph{binary}
if every variable has two values. If all variables have finite numbers of
values, the probability function associated with each variable $v$ can be
represented as a \emph{conditional probability table} (CPT), i.e., a table
with a row for each combination of values that $v$ and its parent vertices can
take. Each row then also has a \emph{probability}, i.e., a number in $[0, 1]$.

WMC is a well-established technique for Bayesian network inference, particularly
effective on networks where most variables have only a few possible values
\cite{DBLP:conf/kr/Darwiche02}. Many ways of encoding a Bayesian network into a
WMC instance have been proposed. We will refer to them based on initials of the
authors and the year of publication. Darwiche was the first to suggest the
\texttt{d02} \cite{DBLP:conf/kr/Darwiche02} encoding that, in many ways, remains
the foundation behind most other encodings. He also introduced the distinction
between \emph{indicator} and \emph{parameter variables}; the former represent
variable-value pairs in the Bayesian network, while the latter are associated
with probabilities in the CPTs. The encoding \texttt{sbk05}
\cite{DBLP:conf/aaai/SangBK05} is the only encoding that deviates from this
arrangement: for each variable in the Bayesian network, one indicator variable
acts simultaneously as a parameter variable. Chavira and Darwiche propose
\texttt{cd05} \cite{DBLP:conf/ijcai/ChaviraD05} where they shift from WMC to
minimum-cardinality WMC because that allows the encoding to have fewer variables
and clauses. In particular, they propose a way to use the same parameter
variable to represent all probabilities in a CPT that are equal and keep only
clauses that `imply' parameter variables (i.e., omit clauses where a parameter
variable implies indicator variables). In their next encoding, \texttt{cd06}
\cite{DBLP:conf/sat/ChaviraD06}, the same authors optimise the aforementioned
implication clauses, choosing the smallest sufficient selection of indicator
variables. A decade later, Bart et al. present \texttt{bklm16}
\cite{DBLP:conf/ecai/BartKLM16} that improves upon \texttt{cd06} in two ways.
First, they optimise the number of indicator variables used per Bayesian network
variable from a linear to a logarithmic amount. Second, they introduce a scaling
factor that can `absorb' one probability per Bayesian network variable. However,
for this work, we choose to disable the latter improvement since this scaling
factor is often small enough to be indistinguishable from zero without the use
of arbitrary precision arithmetic. Indeed, even a small Bayesian network with
seven mutually independent binary variables, 0.1 and 0.9 probabilities each, is
already big enough for the scaling factor to be exactly equal to zero (as
produced by the \texttt{bklm16}
encoder\footnote{\url{http://www.cril.univ-artois.fr/kc/bn2cnf.html}}).

% TODO: incorporate this somewhere
%This additional condition on model
%cardinality becomes necessary because these encodings eliminate clauses of the
%form $p \Rightarrow i$, where $p \in X_P$ is a parameter variable, and $i \in
%X_I$ is an indicator variable. Nonetheless, our transformation algorithm still
%works on such encodings, although the experimental results are discouraging
%because they use approximately twice as many indicator variables. For
%instance, each binary variable of a Bayesian network is encoded using two
%indicator variables while one would suffice.

@@

Below we list some properties of the operations on pseudo-Boolean functions
discussed in this section that can be conveniently represented using our syntax.
The proofs of all these properties follow directly from the definitions.

@@

\section{PBP as a Computational Problem}

@@

\begin{remark}
  Adding scaling factor $\omega$ to the definition allows us to remove
  clauses that consist entirely of a single parameter variable. The idea of
  extracting some of the structure of the WMC instance into an external
  multiplicative factor originates in one of the Bayesian network encodings
  \cite{DBLP:conf/ecai/BartKLM16}.
\end{remark}

\subsection{From WMC to PBP}

\paragraph{Notes.}
\begin{itemize}
\item The algorithm is based on several observations that will be made precise
  in \cref{thm:correctness}. First, all weights except for $\{w(p) \mid p \in
  X_P\}$ are redundant as they either duplicate an already-defined weight or are
  equal to one. Second, each clause has at most one parameter variable. Third,
  if the parameter variable is negated, we can ignore the clause (this idea
  first appears in the \texttt{cd05} paper \cite{DBLP:conf/ijcai/ChaviraD05}).
\item We formulate our algorithm as a sequel to the WMC encoding procedure
  because the implementations of Bayesian network WMC encodings are all
  closed-source. However, as all transformations in \cref{alg:transformation}
  are local, the algorithm can be efficiently incorporated into a WMC encoding
  algorithm with no noticeable slowdown.
\item No obvious way to do this for \texttt{sbk05} because the roles of
  indicator and parameter (i.e., `chance') variables overlap
  \cite{DBLP:conf/aaai/SangBK05}.
\item Benefits of having this proof in the paper:
  \begin{itemize}
  \item It puts all encodings on a common ground.
  \item It illustrates the convenience of our notation for reasoning about
    (certain types of) pseudo-Boolean functions.
  \item It's too big and too important to be left for the appendix.
  \end{itemize}
\end{itemize}

%\item The second \textbf{foreach} loop can be performed in constant time by
%  representing $\phi'$ as a list and assuming that the two 'clauses' are
%  adjacent in that list (and incorporating it into the first loop).
%\item The $d$ map is constructed in $\mathcal{O}(|X_P|\log|X_P|)$ time (we want
%  to use a data structure based on binary search trees rather than hashing).
%\item \texttt{rename} can be implemented in $\mathcal{O}(\log |X_P|)$ time.

\begin{algorithm}[t]
  \caption{WMC to PBP transformation}
  \label{alg:transformation}
  \SetKwFunction{rename}{rename}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{WMC instance $(\phi, X_I, X_P, w)$}
  \KwResult{PBP instance $(F, X, \omega)$}
  $F \gets \emptyset$\;
  $\omega \gets 1$\;
  let $d\colon X_P \to \mathbb{N}$ be defined as $p \mapsto |\{ o \in X_P \mid o
  \le p \}|$\;
  \ForEach{clause $c \in \phi$\label{line:foreach1start}}{
    \uIf{$c \cap X_P = \{ p \}$ for some $p$ \textnormal{\textbf{and}} $w(p) \ne
      1$}{
      \uIf{$|c| = 1$}{$\omega \gets \omega \times w(p)$\;}
      \Else{
        $F \gets F \cup \left\{ \left[ \bigwedge_{l \in c \setminus
              \{p\}} \neg l \right]^{w(p)}_1 \right\}$\;
      }
    }
    \ElseIf{$\{p \mid \neg p \in c\} \cap X_P  = \emptyset$}{
      $F \gets F \cup \{ [c]^1_0 \}$\; \label{line:foreach1end}
    }
  }
  \ForEach{indicator variable $v \in X_I$ \label{line:foreach2start}}{
    \If{$\{[v]_1^p, [\neg v]_1^q\} \subseteq F$ for some $p$ and $q$}{
      $F \gets F \setminus \{ [v]_1^p, [\neg v]_1^q \} \cup \{ [v]_q^p
      \}$\; \label{line:foreach2end}
    }
  }
  replace every variable $v$ in $F$ with $\rename{$v$}$\;
  \Return{$(F, X_I, \omega)$}\;
  \Fn{\rename{$v$}}{
    $S \gets \{u \in X_P \mid u \le v\}$\;
    \lIf{$S = \emptyset$}{\Return{$v$}}
    \Return{$v - d(\max S)$}\;
  }
\end{algorithm}

@@

\paragraph{Notes.}
\begin{itemize}
\item This just says that $p$ is equivalent to a conjunction.
\item The first group of conditions applies to \texttt{d02}, while the second
  group applies to \texttt{bklm16}.
\item For \texttt{cd05} and \texttt{cd06}, condition $2bi$ should be replaced
  with $w(\neg p) = 1$.
\end{itemize}

\paragraph{TODO}
\begin{itemize}
\item check if each condition is actually used. Maybe turn this into a paragraph
  that gives an overview of the proof.
\item condition 1 is necessary in both cases because we're ignoring the weights
  of indicator variables (explicitly acknowledge this)
\item add the \texttt{cd05}/\texttt{cd06} correctness theorem
\end{itemize}

@@

% TODO: mention that this is similar (but not equal) to the DPMC/ADDMC result
\begin{proposition} \label{prop:equivalence}
  Let $(\phi, X_I, X_P, w)$ be a WMC instance. Then
  \begin{equation}
  \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
        x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
    1\right) \label{eq:new_wmc}
  \end{equation}
  is a PBP instance with the same answer.
\end{proposition}
\begin{proof}
  Let $f = \prod_{c \in \phi} [c]_0^1$, and $g = \prod_{x \in X_I \cup X_P}
  [x]_{w(\neg x)}^{w(x)}$. Then the answer to the WMC instance
  \eqref{eq:new_wmc} is
  \[
    (\exists_{X_I \cup X_P} fg)(\emptyset) = \sum_{Y \subseteq X_I \cup X_P}
    (fg)(Y) = \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y)
  \]
  by \cref{lemma:sum}. Note that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that
  \[
    \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models \phi} \prod_{Y
      \models l} w(l)
  \]
  as required.
\end{proof}

\begin{theorem}[Correctness] \label{thm:correctness}
  \Cref{alg:transformation}, when given a WMC instance $(\phi, X_I, X_P,
  w)$, returns PBP instance with the same answer, provided the following
  conditions are satisfied:
  \begin{enumerate}
  \item for all indicator variables $i \in X_I$, $w(i) = w(\neg i) = 1$,
  \item and either
    \begin{enumerate}
    \item for all parameter variables $p \in X_P$, there is a non-empty family
      of literals $(l_i)_{i=1}^n$ such that \label{condition:d02}
      \begin{enumerate}
      \item $w(\neg p) = 1$,
      \item $l_i \in X_I$ or $\neg l_i \in X_I$ for all $i = 1, \dots, n$,
        \label{condition:equivalence1}
      \item and $\{ c \in \phi \mid p \in c \text{ or } \neg p \in c \} =
        \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
        \mid i = 1, \dots, n \}$; \label{condition:equivalence2}
      \end{enumerate}
    \item or for all parameter variables $p \in X_P$, \label{condition:bklm16}
      \begin{enumerate}
      \item $w(p) + w(\neg p) = 1$,
      \item for any clause $c \in \phi$, $|c \cap X_P| \le 1$,
      \item there is no clause $c \in \phi$ such that $\neg p \in c$,
      \item if $\{ p \} \in \phi$, then there is no clause $c \in \phi$ such
        that $c \ne \{ p \}$ and $p \in c$, \label{cond:just_parameter}
        % 'prime implicants are logically disjoint'
      \item and for any $c, d \in \phi$ such that $c \ne d$, $p \in c$ and $p
        \in d$, $\bigwedge_{l \in c \setminus \{ p \}} \neg l \land \bigwedge_{l
          \in d \setminus \{ p \}} \neg l$ is false. \label{cond:disjoint}
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}
\end{theorem}
\begin{proof}
  By \cref{prop:equivalence},
  \begin{equation}
    \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
          x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
      1\right) \label{eq:new_wmc2}
  \end{equation}
  is a PBP instance with the same answer as the given WMC instance. By
  \cref{def:new_wmc}, its answer is
  \begin{equation} \label{eq:answer}
    \left(\exists_{X_I \cup X_P} \left( \prod_{c \in \phi} [c]_0^1 \right) \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset)
  \end{equation}
  Since both \cref{condition:d02,condition:bklm16} ensure that each clause in
  $\phi$ has at most one parameter variable, we can partition $\phi$ into
  $\phi_* \coloneqq \{c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \emptyset \}$
  and $\phi_p \coloneqq \{ c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \{ p \}
  \}$ for all $p \in X_P$. We can then use \cref{thm:early} to reorder
  \eqref{eq:answer} into
  \[
    \left(\exists_{X_I} \left( \prod_{x \in X_I} [x]_{w(\neg x)}^{w(x)} \right)
      \left( \prod_{c \in \phi_*} [c]_0^1 \right) \prod_{p \in X_P} \exists_p
      [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 \right)(\emptyset).
  \]

@@

\section{Experimental Evaluation}
% TODO: what are the encodings compiled with?

We run a set of experiments, comparing all five original Bayesian network
encodings (\texttt{bklm16}, \texttt{cd05}, \texttt{cd06}, \texttt{d02}
\texttt{sbk05}) as well as the first four with \cref{alg:transformation} applied
afterwards.\footnote{Recall that \texttt{cd05} and \texttt{cd06} are
  incompatible with \textsf{DPMC}.} For each encoding \texttt{e}, we will write
\texttt{e++} to denote the combination of encoding a Bayesian network as a WMC
instance using \texttt{e} and transforming it into a PBP instance using
\cref{alg:transformation}. Along with
\textsf{DPMC}\footnote{\url{https://github.com/vardigroup/DPMC}}, we also
include WMC algorithms used in the papers that introduce each encoding:
\textsf{Ace} for \texttt{cd05}, \texttt{cd06}, and \texttt{d02};
\textsf{Cachet}\footnote{\url{https://cs.rochester.edu/u/kautz/Cachet/}}
\cite{DBLP:conf/sat/SangBBKP04} for \texttt{sbk05}; and
\textsf{c2d}\footnote{\url{http://reasoning.cs.ucla.edu/c2d/}}
\cite{DBLP:conf/ecai/Darwiche04} with
\textsf{query-dnnf}\footnote{\url{http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html}}
for \texttt{bklm16}. We focus on the following questions:
\begin{itemize}
\item Can parameter variable elimination improve inference speed?
\item How does DPMC combined with encodings without (and with) parameter
  variables compare with other WMC algorithms and other encodings?
\item Which instances is our approach particularly successful on (compared to
  other algorithms and encodings and to the same encoding before our
  transformation)?
\item What proportion of variables is typically removed?
\item Do some encodings benefit from this transformation more than others?
\end{itemize}

\subsection{Setup (mostly done)}

\textsf{DPMC} is run with tree decomposition based planning and algebraic
decision diagram based execution---the best-performing combination in the
original set of experiments \cite{DBLP:conf/cp/DudekPV20}. We use a single
iteration of \textsf{htd} \cite{DBLP:conf/cpaior/AbseherMW17} to generate
approximately optimal tree decompositions---we found that this configuration
is efficient enough to handle huge instances, and yet the width of the
returned decomposition is unlikely to differ from optimal by more than one or
two. We also enabled \textsf{DPMC}'s greedy mode. This mode (which was not part
of the original paper \cite{DBLP:conf/cp/DudekPV20}) optimises the order in
which pseudo-Boolean functions are multiplied by prioritising functions with
small representations.

For experimental data, we use Bayesian networks available with \textsf{Ace} and
\textsf{Cachet}. We split them into the following groups:
\begin{itemize}
\item DQMR (390 instances) and
\item Grid networks (450 instances) as described by Sang et al.
  \cite{DBLP:conf/aaai/SangBK05};
\item Mastermind (144 instances) and
\item Random Blocks (256 instances) by Chavira et al.
  \cite{DBLP:journals/ijar/ChaviraDJ06};
\item other binary Bayesian networks (50 instances) including Plan Recognition
  \cite{DBLP:conf/aaai/SangBK05}, Friends and Smokers, Students and Professors
  \cite{DBLP:journals/ijar/ChaviraDJ06}, and \texttt{tcc4f};
\item non-binary classic networks (176 instances): \texttt{alarm},
  \texttt{diabetes}, \texttt{hailfinder}, \texttt{mildew},
  \texttt{munin1}--\texttt{4}, \texttt{pathfinder}, \texttt{pigs}, and
  \texttt{water}.
\end{itemize}

To perform Bayesian network inference with \textsf{DPMC} (or with any
other WMC algorithm not based on compilation such as \textsf{Cachet}), one
needs to select a probability to compute
\cite{DBLP:conf/cp/DudekPV20,DBLP:conf/sat/SangBBKP04}. If a network comes
with an evidence file, we compute the probability of this evidence. Otherwise,
let $X$ be the variable last mentioned in the Bayesian network file. If
\textsf{true} is one of the values of $X$, then we compute $\Pr(X =
\textsf{true})$, otherwise we choose the first-mentioned value of $X$.
